---
title: "Galapagos Blacktip Diet"
author: "Savannah Ryburn"
date: "2025-05-21"
output: html_document
---

Loading packages
```{r}
library(ape)
library(phyloseq)
library(ggplot2)
library(tibble)
library(dplyr)
library(tidyverse)
library(vegan)
library(tidyr)
library(car)
library(rstatix)
library(devtools)
library(pairwiseAdonis)
library(spaa)
library(dietr)
library(ggridges)
library(microViz)
library(ggraph)
library(DT)
library(corncob)
library(metagMisc)
```

Load phyloseq object of just fecal swabs made in script 10
```{r}
Gal_Shark_Diet.ps<-readRDS(file = "/Users/savannahryburn/Desktop/Galapagos_Sharks/10_Phyloseq/Gal_Shark_Diet.ps.RDS")

Gal_Shark_Diet.ps
```

Combine OTUs from the two fish species that weren't present in the Galapagos with the lowest taxonomic group
```{r}
# New phyloseq object so I don't mess up the old one
Gal_Shark_Diet.ps.new1 <- Gal_Shark_Diet.ps

# Extract matrix safely
otu_mat <- as(otu_table(Gal_Shark_Diet.ps.new1), "matrix")

# Double-check the OTU names (Make sure this returns exactly: "Abudefduf genus 1"  "Abudefduf taurus 1")
grep("Abudefduf", rownames(otu_mat), value = TRUE)

# save them explicitly:
otu1 <- "Abudefduf genus 1"
otu2 <- "Abudefduf taurus 1"

# Sum the two OTUs into a new named row
# Check if both OTUs exist
if (all(c(otu1, otu2) %in% rownames(otu_mat))) {
  new_otu_row <- otu_mat[otu1, ] + otu_mat[otu2, ]
  # Add the new row
  otu_mat <- rbind(otu_mat, "Abudefduf" = new_otu_row)
  # Remove the original two
  otu_mat <- otu_mat[!rownames(otu_mat) %in% c(otu1, otu2), ]
} else {
  stop("One or both OTU names not found!")
}

nrow(otu_mat)  # should be 32

# Rebuild OTU table (make sure it has the right orientation)
otu_tab_new <- otu_table(otu_mat, taxa_are_rows = TRUE)

# Replace the OTU table in your phyloseq object
Gal_Shark_Diet.ps.new1 <- phyloseq(
  otu_tab_new,
  tax_table(Gal_Shark_Diet.ps.new1),
  sample_data(Gal_Shark_Diet.ps.new1))

dim(otu_table(Gal_Shark_Diet.ps.new1))  # Should show 32 rows
rownames(otu_table(Gal_Shark_Diet.ps.new1))  # Should include "Abudefduf"




head(tax_table(Gal_Shark_Diet.ps.new1))

# Manually create new row
# Extract the existing tax table
tax_mat <- as(tax_table(Gal_Shark_Diet.ps.new1), "matrix")

# Create a new row (fill in with correct levels from your data structure)
new_tax_row <- c(
  Kingdom = "Eukaryota",
  Phylum = "Chordata",
  Class = "Actinopterygii",
  Order = "Perciformes",
  Family = "Pomacentridae",
  Genus = "Abudefduf",
  Species = "Abudefduf genus"  # or "Abudefduf spp." or leave NA
)

# Add the new row with the OTU name "Abudefduf"
tax_mat <- rbind(tax_mat, "Abudefduf" = new_tax_row)

# Recreate tax_table object
new_tax_table <- tax_table(tax_mat)

# Rebuild phyloseq object
Gal_Shark_Diet.ps.new1 <- phyloseq(
  otu_tab_new,
  new_tax_table,
  sample_data(Gal_Shark_Diet.ps.new1)
)

dim(otu_table(Gal_Shark_Diet.ps.new1))       # Should now be 32 x 172
rownames(otu_table(Gal_Shark_Diet.ps.new1))  # Should include "Abudefduf"
tax_table(Gal_Shark_Diet.ps.new1)["Abudefduf", ]  # Should show your new taxonomy

View(Gal_Shark_Diet.ps.new1)
View(Gal_Shark_Diet.ps.new1@tax_table)



####### Now do this again for "Cyprinodon variegatus 1" and "Actinopteri class 1"

# save them explicitly:
otu3 <- "Cyprinodon variegatus 1"
otu4 <- "Actinopteri class 1"

# Sum the two OTUs into a new named row
# Check if both OTUs exist
if (all(c(otu3, otu4) %in% rownames(otu_mat))) {
  new_otu_row <- otu_mat[otu3, ] + otu_mat[otu4, ]
  # Add the new row
  otu_mat <- rbind(otu_mat, "Actinopteri" = new_otu_row)
  # Remove the original two
  otu_mat <- otu_mat[!rownames(otu_mat) %in% c(otu3, otu4), ]
} else {
  stop("One or both OTU names not found!")
}

nrow(otu_mat)  # should be 31

# Rebuild OTU table (make sure it has the right orientation)
otu_tab_new <- otu_table(otu_mat, taxa_are_rows = TRUE)

# Replace the OTU table in your phyloseq object
Gal_Shark_Diet.ps.new1 <- phyloseq(
  otu_tab_new,
  tax_table(Gal_Shark_Diet.ps.new1),
  sample_data(Gal_Shark_Diet.ps.new1)
)

dim(otu_table(Gal_Shark_Diet.ps.new1))  # Should show 32 rows
rownames(otu_table(Gal_Shark_Diet.ps.new1))  # Should include "Abudefduf"


head(tax_table(Gal_Shark_Diet.ps.new1))

# Manually create new row
# Extract the existing tax table
tax_mat <- as(tax_table(Gal_Shark_Diet.ps.new1), "matrix")

# Create a new row (fill in with correct levels from your data structure)
new_tax_row <- c(
  Kingdom = "Eukaryota",
  Phylum = "Chordata",
  Class = "Actinopteri",
  Order = "Actinopteri class",
  Family = "Actinopteri class",
  Genus = "Actinopteri class",
  Species = "Actinopteri class")

# Add the new row with the OTU name "Actinopteri"
tax_mat <- rbind(tax_mat, "Actinopteri" = new_tax_row)

# Recreate tax_table object
new_tax_table <- tax_table(tax_mat)

# Rebuild phyloseq object
Gal_Shark_Diet.ps.new1 <- phyloseq(
  otu_tab_new,
  new_tax_table,
  sample_data(Gal_Shark_Diet.ps.new1)
)

dim(otu_table(Gal_Shark_Diet.ps.new1))       # Should now be 31 x 172
rownames(otu_table(Gal_Shark_Diet.ps.new1))  # Should include "Actinopteri"
tax_table(Gal_Shark_Diet.ps.new1)["Actinopteri", ]  # Should show your new taxonomy

View(Gal_Shark_Diet.ps.new1)
View(Gal_Shark_Diet.ps.new1@tax_table)
View(Gal_Shark_Diet.ps.new1@otu_table)
```

Divide blacktips and scalloped hammerheads
```{r}
Gal_Shark_Diet_BT.ps<-subset_samples(Gal_Shark_Diet.ps.new1,Species..scientific.name.=="Carcharhinus limbatus")

View(Gal_Shark_Diet_BT.ps@otu_table)
View(Gal_Shark_Diet_BT.ps@tax_table)
```

Filter out host DNA (in this case blacktip DNA)
```{r}
tax_table(Gal_Shark_Diet_BT.ps)

# Exclude "Carcharhinus genus"
Gal_Shark_Diet_BT.ps <- subset_taxa(Gal_Shark_Diet_BT.ps, species != "Carcharhinus genus")

View(Gal_Shark_Diet_BT.ps@otu_table)
View(Gal_Shark_Diet_BT.ps@tax_table)
```

Add new columns to data sheet (e.g., trophic levels) - don't need this until I look up all of the extra info
```{r}
##for prey data (tax_table)
newtaxa.df<-as.data.frame(cbind(tax_table(Gal_Shark_Diet_BT.ps)))

#write.csv(newtaxa.df, "/Users/savannahryburn/Desktop/Chapter3/Galapagos_Blacktip_Diet/taxafiletoedit.csv")

##now edit the file and load it in using this
  #(I changed this from row.names = 1 to NULL so that the first column wouldn't be a header - make sure this didn't mess anything else up, I did this for the trophic level calculations)
editedtaxa.df <- read.csv("/Users/savannahryburn/Desktop/Chapter3/Galapagos_Blacktip_Diet/editedtaxafile.csv", header = FALSE, row.names = 1)

names(editedtaxa.df) <- lapply(editedtaxa.df[1, ], as.character)
editedtaxa.df <- editedtaxa.df[-1,]
View(editedtaxa.df)


##for shark data (sample_data)
newsample.df<-as.data.frame(cbind(sample_data(Gal_Shark_Diet.ps)))

#write.csv(newsample.df, "/Users/savannahryburn/Desktop/Chapter3/Galapagos_Blacktip_Diet/samplefiletoedit.csv")

##now edit the file and load it in using this
#editedsample.df <- read.csv("/Users/savannahryburn/Desktop/Chapter3/Galapagos_Blacktip_Diet/editedsamplefile.csv", row.names = "X")
editedsample.df <- read.csv("/Users/savannahryburn/Desktop/Chapter3/Galapagos_Blacktip_Diet/editedsamplefile.csv", row.names = 1)
View(editedsample.df)


##Prune scalloped hammerheads, samples with 0 reads, and the 3 control samples
# Subset the data frame for blacktip shark samples
editedsample.df <- editedsample.df %>%
  filter(Species..scientific.name. == "Carcharhinus limbatus")

# Specify the sample names to remove
samples_to_remove <- c("con_G15", "con_G31", "con_G43", "G105", "G106", "G109", "G116", "G117", "G121", "G125", "G127", "G128", "G129", "G134", "G163", "G166", "G190")

# Remove the specified samples
editedsample.df <- editedsample.df %>%
  filter(!(rownames(.) %in% samples_to_remove))

# View the updated data frame
View(editedsample.df)

#make new phyloseq object so I don't mess up th original
Gal_Shark_Diet_edited.ps<-Gal_Shark_Diet_BT.ps

sample_data(Gal_Shark_Diet_edited.ps)<-editedsample.df
#tax_table(Gal_Shark_Diet_edited.ps)<-editedtaxa.df
tax_table(Gal_Shark_Diet_edited.ps) <- as.matrix(editedtaxa.df)

sample_variables(Gal_Shark_Diet_edited.ps)

View(Gal_Shark_Diet_edited.ps)

##Now prune the taxa that are only present in blacktips
# Exclude "Carcharhinus genus"
Gal_Shark_Diet_edited.ps <- subset_taxa(Gal_Shark_Diet_edited.ps, species != "Carcharhinus genus")
# Remove taxa with no reads
Gal_Shark_Diet_edited.ps <- prune_taxa(taxa_sums(Gal_Shark_Diet_edited.ps) > 0, Gal_Shark_Diet_edited.ps)
# View the updated phyloseq object
Gal_Shark_Diet_edited.ps

#merge motus by prey species
Gal_Shark_Diet_edited.ps_prey_species_merge.ps <- tax_glom(Gal_Shark_Diet_edited.ps, taxrank = "species", NArm = FALSE)

# Removing the NAs (anything that wasn't IDed to species gets deleted here)
Gal_Shark_Diet_edited.ps_prey_species_merge.ps.1<-tax_glom(Gal_Shark_Diet_edited.ps_prey_species_merge.ps, taxrank = "species", NArm = TRUE)
 rank_names(Gal_Shark_Diet_edited.ps_prey_species_merge.ps.1)

Gal_Shark_Diet_edited.ps %>%
  tax_fix(
   min_length = 4,
   unknowns = c(""),
   sep = "", anon_unique = TRUE,
   suffix_rank = "classified")

View(Gal_Shark_Diet_edited.ps@otu_table)
```

Make a Final Taxa Table with the number of reads assigned to each genus or species
```{r}
#merge motus by prey species - I don't think I need this because I did this in script 10
#NC_Shark_prey_species_merge.ps <- tax_glom(NC_Sharks1.ps, taxrank = "species", NArm = FALSE)

sample_variables(Gal_Shark_Diet_BT.ps)
rank_names(Gal_Shark_Diet_BT.ps)

#export read table with species or genus (if only genus available) instead of motu/asv now that they've been properly merged
taxtbl <- as.data.frame(tax_table(Gal_Shark_Diet_BT.ps))
otutbl<-as.data.frame(otu_table(Gal_Shark_Diet_BT.ps))
sampletbl<-as.data.frame(sample_data(Gal_Shark_Diet_BT.ps))

otutbl$species <-
  taxtbl$species[match(rownames(otutbl),rownames(taxtbl))]

otutbl$genus <-
  taxtbl$genus[match(rownames(otutbl),rownames(taxtbl))]

df<-otutbl %>% relocate(genus)
df<-df %>% relocate(species, .after = genus)

df1<- as.data.frame(t(df))

df1<- rownames_to_column(df1, var="sample")
samples.df.1<-rownames_to_column(sampletbl, var="sample")
df2<- left_join(df1, samples.df.1, by ="sample")

colnames(df2)

df3<- df2 %>% select(-"type",-"control_type",-"nb_reads.x",-"nb_motus.x",-"seqdepth_ok.x",-"nb_reads_postmetabaR.x",-"nb_motus_postmetabaR.x",-"nb_reads.y",-"nb_motus.y",-"seqdepth_ok.y",-"nb_reads_postmetabaR.y",-"nb_motus_postmetabaR.y")

colnames(df3)

write.csv(df3, "Gal_BT_samples_by_species_combined_table.csv")

#Flip axis of the data frame
transformed_df3 <- t(df3)
write.csv(transformed_df3, "Gal_BT_samples_by_species_combined_table_flipped.csv")

################################################################################################################

#Make table that includes family and order too
otutbl$species <-
  taxtbl$species[match(rownames(otutbl),rownames(taxtbl))]

otutbl$genus <-
  taxtbl$genus[match(rownames(otutbl),rownames(taxtbl))]

otutbl$family <-
  taxtbl$family[match(rownames(otutbl),rownames(taxtbl))]

otutbl$order <-
  taxtbl$order[match(rownames(otutbl),rownames(taxtbl))]

df<-otutbl %>% relocate(species) %>% relocate(genus) %>% relocate(family) %>% relocate(order)
#df<-df %>% relocate(species, .after = genus)

df1<- as.data.frame(t(df))

df1<- rownames_to_column(df1, var="sample")
samples.df.1<-rownames_to_column(sampletbl, var="sample")
df2<- left_join(df1, samples.df.1, by ="sample")

colnames(df2)

df3<- df2 %>% select(-"type",-"control_type",-"nb_reads.x",-"nb_motus.x",-"seqdepth_ok.x",-"nb_reads_postmetabaR.x",-"nb_motus_postmetabaR.x",-"nb_reads.y",-"nb_motus.y",-"seqdepth_ok.y",-"nb_reads_postmetabaR.y",-"nb_motus_postmetabaR.y")

colnames(df3)

write.csv(df3, "Gal_BT_samples_by_species_combined_table_with familyand order.csv")

edited_df3 <- read.csv("/Users/savannahryburn/Desktop/Chapter3/Galapagos_Blacktip_Diet/Gal_BT_samples_by_species_combined_table_with familyand order_edited.csv")
View(edited_df3)

#Flip axis of the data frame
transformed_edited_df3 <- t(edited_df3)
write.csv(transformed_edited_df3, "Gal_BT_samples_by_species_combined_table_with familyand order_edited_flipped.csv")
```

Prune taxa with zero abundance across all samples and get rid of controls
```{r}
# Check the sample sums (total reads per sample)
sample_sums(Gal_Shark_Diet_BT.ps)
# Remove samples with zero reads
Gal_Shark_Diet_BT.ps <- prune_samples(sample_sums(Gal_Shark_Diet_BT.ps) > 0, Gal_Shark_Diet_BT.ps)
# Verify that the samples with zero reads are removed
sample_sums(Gal_Shark_Diet_BT.ps)

# List the names of the control samples to remove
controls <- c("con_G15", "con_G31", "con_G43")
# Remove the controls
Gal_Shark_Diet_BT.ps <- prune_samples(!(sample_names(Gal_Shark_Diet_BT.ps) %in% controls), Gal_Shark_Diet_BT.ps)
# Verify the controls are removed
sample_names(Gal_Shark_Diet_BT.ps)

##Now prune the taxa that are only present in scalloped hammerheads and not blacktips
# Remove taxa with no reads
Gal_Shark_Diet_BT.ps <- prune_taxa(taxa_sums(Gal_Shark_Diet_BT.ps) > 0, Gal_Shark_Diet_BT.ps)
# View the updated phyloseq object
Gal_Shark_Diet_BT.ps

View(Gal_Shark_Diet_BT.ps@otu_table)
```

POO percentages with all sharks combined
```{r}
#Convert to presence absence: (FOO)
Gal_Shark_Diet_BT.ps.PA <- phyloseq_standardize_otu_abundance(Gal_Shark_Diet_BT.ps, method = "pa")

# Prune taxa with zero abundance across all samples
Gal_Shark_Diet_BT.ps.PA <- prune_taxa(
  taxa_sums(Gal_Shark_Diet_BT.ps.PA) > 0, 
  Gal_Shark_Diet_BT.ps.PA)

#Put it into dataframes
otu_df<-cbind(data.frame(otu_table(Gal_Shark_Diet_BT.ps.PA)))
otu_df$otu<-rownames(otu_df)

otu_df_long<-otu_df%>%
  pivot_longer(cols = -otu,names_to = "sample",values_to = "presence")

otu_df_long$sample <-gsub(pattern = ".",
                          "-",
                          x=otu_df_long$sample,
                          fixed = TRUE)

sample_df<-cbind(data.frame(sample_data(Gal_Shark_Diet_BT.ps.PA)))
sample_df<-rownames_to_column(sample_df, var = "sample")
sample_site_key<-sample_df%>%
  select(sample,Species..scientific.name.)

taxa_df<-cbind(data.frame(tax_table(Gal_Shark_Diet_BT.ps.PA)))
taxa_df<-rownames_to_column(taxa_df,var="otu")
otu_species_key<-taxa_df%>%
  select(otu,family,genus,species)

otu_df_long<-otu_df_long%>%
  left_join(sample_site_key)%>%
  left_join(otu_species_key)

#group by different taxonomic levels (shark species)

merge_shark.species_otu_counts<-otu_df_long%>%
  group_by(Species..scientific.name.,species)%>%
  summarise(total_count=sum(presence))%>%
  ungroup()

merge_shark.species_counts<-merge_shark.species_otu_counts%>%
  group_by(Species..scientific.name.)%>%
  mutate(total_site=sum(total_count))%>%
  ungroup()%>%
  mutate(rel_prct=(total_count/total_site)*100)

View(merge_shark.species_counts)
```

POO Plot separating blacktips by site
```{r}
#Convert to presence absence: (FOO)
Gal_Shark_Diet_BT.ps.PA <- phyloseq_standardize_otu_abundance(Gal_Shark_Diet_BT.ps, method = "pa")

# Prune taxa with zero abundance across all samples
Gal_Shark_Diet_BT.ps.PA <- prune_taxa(
  taxa_sums(Gal_Shark_Diet_BT.ps.PA) > 0, 
  Gal_Shark_Diet_BT.ps.PA)

#Top 20
# top20occ = names(sort(taxa_sums(Gal_Shark_Diet_BT.ps.PA), decreasing = TRUE)[1:20])
# Gal_Shark_Diet_BT.ps.PA.Top20 = prune_taxa(top20occ, Gal_Shark_Diet_BT.ps.PA.Top20)

#Put it into dataframes
otu_df<-cbind(data.frame(otu_table(Gal_Shark_Diet_BT.ps.PA)))
otu_df$otu<-rownames(otu_df)

otu_df_long<-otu_df%>%
  pivot_longer(cols = -otu,names_to = "sample",values_to = "presence")

otu_df_long$sample <-gsub(pattern = ".",
                          "-",
                          x=otu_df_long$sample,
                          fixed = TRUE)

sample_df<-cbind(data.frame(sample_data(Gal_Shark_Diet_BT.ps.PA)))
sample_df<-rownames_to_column(sample_df, var = "sample")
sample_site_key<-sample_df%>%
  select(sample,Site.Name)

taxa_df<-cbind(data.frame(tax_table(Gal_Shark_Diet_BT.ps.PA)))
taxa_df<-rownames_to_column(taxa_df,var="otu")
otu_species_key<-taxa_df%>%
  select(otu,family,genus,species)

otu_df_long<-otu_df_long%>%
  left_join(sample_site_key)%>%
  left_join(otu_species_key)

#group by different taxonomic levels (shark species)

merge_site_otu_counts<-otu_df_long%>%
  group_by(Site.Name,species)%>%
  summarise(total_count=sum(presence))%>%
  ungroup()

merge_site_counts<-merge_site_otu_counts%>%
  group_by(Site.Name)%>%
  mutate(total_site=sum(total_count))%>%
  ungroup()%>%
  mutate(rel_prct=(total_count/total_site)*100)

View(merge_site_counts)

#plot 
POO_BT_site_plot <- ggplot(merge_site_counts, fill = "species") +
  geom_bar(aes(x = Site.Name, y = rel_prct, fill = species), 
           stat = "identity", position = "stack", color = NA) + # Remove borders
  ylab("Relative Percent of Occurrence") + 
  xlab("Shark Nursery Bay") +
  theme_classic() + 
  theme(text = element_text(size = 11)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "bottom",             # Move legend below the graph
        legend.box = "horizontal",              # Arrange legend items horizontally
        legend.box.spacing = unit(0.4, "cm")) + 
  scale_fill_manual(
    values = c(
      "Abudefduf genus" = "lightblue", 
      "Actinopteri class" = "navyblue", "Amphipholis squamata" = "hotpink1", 
      "Anisotremus scapularis" = "palegreen", "Caranx crysos" = "royalblue", 
      "Decapterus genus" = "indianred4", 
      "Euthynnus affinis" = "slategray2", "Fistularia commersonii" = "khaki", 
      "Grapsus grapsus" = "tomato", "Hemiramphus lutkei" = "green3", 
      "Lutjanus genus" = "aquamarine4", "Mugil cephalus" = "blue", 
      "Mugil thoburni" = "purple", "Muraena lentiginosa" = "lightpink", 
      "Pachygrapsus socius" = "goldenrod3", "Sardinops genus" = "red", 
      "Scarus ghobban" = "yellow", "Selar crumenophthalmus" = "slateblue1", 
      "Semicossyphus pulcher" = "lawngreen", "Sphoeroides annulatus" = "lightseagreen", 
      "Sphyrna lewini" = "yellowgreen", "Stegastes arcifrons" = "darkgreen", 
      "Stegastes beebei" = "orange", "Teleophrys cristulipes" = "turquoise2", 
      "Thunnus genus" = "darksalmon"),
    labels = c(
      "Abudefduf genus" = "Abudefduf spp. (sergeant-majors)",
      "Actinopteri class" = "Actinopteri (ray-finned fish)",
      "Amphipholis squamata" = "Amphipholis squamata (dwarf brittle star)",
      "Anisotremus scapularis" = "Anisotremus scapularis (Peruvian grunt)",
      "Caranx crysos" = "Caranx crysos (blue runner)",
      "Decapterus genus" = "Decapterus spp. (jack)",
      "Euthynnus affinis" = "Euthynnus affinis (mackerel tuna)",
      "Fistularia commersonii" = "Fistularia commersonii (bluespotted conetfish)",
      "Grapsus grapsus" = "Grapsus grapsus (Sally lightfoot crab)",
      "Hemiramphus lutkei" = "Hemiramphus lutkei (Lutke's halfbeak)",
      "Lutjanus genus" = "Lutjanus spp. (snapper)",
      "Mugil cephalus" = "Mugil cephalus (flathead grey mullet)",
      "Mugil thoburni" = "Mugil thoburni (Thoburn's mullet)",
      "Muraena lentiginosa" = "Muraena lentiginosa (jewel eel)",
      "Pachygrapsus socius" = "Pachygrapsus Socius (Pacific mottled shore crab)",
      "Sardinops genus" = "Sardinops spp. (sardine)",
      "Scarus ghobban" = "Scarus ghobban (blue-barred parrotfish)",
      "Selar crumenophthalmus" = "Selar crumenophthalmus (bigeye scad)",
      "Semicossyphus pulcher" = "Semicossyphus pulcher (California sheephead)",
      "Sphoeroides annulatus" = "Sphoeroides annulatus (bullseye puffer)",
      "Sphyrna lewini" = "Sphyrna lewini (scalloped hammerhead shark)",
      "Stegastes arcifrons" = "Stegastes arcifrons (yellowtail damselfish)",
      "Stegastes beebei" = "Stegastes beebei (Galapagos ringtail damselfish)",
      "Teleophrys cristulipes" = "Teleophrys cristulipes (spider crab)",
      "Thunnus genus" = "Thunnus spp. (tuna)")) +
  scale_x_discrete(
    limits = c(
      "Puerto Grande", "La Seca", "Cerro Brujo", "Rosa Blanca 1", 
      "Rosa Blanca 2", "La Tortuga", "Cartago Bay", "Albermarle", 
      "Grimanesa", "Puerto Chino", "Baleado", "El Mulle de Balleno", 
      "La Bomba", "Punta Espinoza"))


POO_BT_site_plot
```

Statistics for MDS plot (prey at genus level) - jsd
```{r}
# Removing the NAs (anything that wasn't IDed to genus gets deleted here)
Gal_Shark_Diet_BT_prey_genus.ps<-tax_glom(Gal_Shark_Diet_BT.ps, taxrank = "genus", NArm = TRUE)

# play with ordinations interactively
pseq_BT_prey_genus <- Gal_Shark_Diet_BT_prey_genus.ps %>%
  phyloseq_validate()

jsd_dist_matrix <- phyloseq::distance(pseq_BT_prey_genus, method ="jsd")

#Between sites
set.seed(200)
permanova <- vegan::adonis2(jsd_dist_matrix ~ phyloseq::sample_data(pseq_BT_prey_genus)$Site.Name)
permanova
View(permanova)

# set.seed(200)
# pairwise.adonis(jsd_dist_matrix, phyloseq::sample_data(pseq_BT_prey_genus)$Site.Name)

#Between islands
set.seed(200)
permanova <- vegan::adonis2(jsd_dist_matrix ~ phyloseq::sample_data(pseq_BT_prey_genus)$Island)
permanova
View(permanova)
```

Rarefaction curve (includes genus and species, double check this)
```{r}
Rarefaction_BT <- read.csv("Rarefaction_Gal_BT.csv")
#View(Rarefaction_BT)

#Switch axis
Rarefaction_BT_Reverse <- data.frame(t(Rarefaction_BT[-1]))
colnames(Rarefaction_BT_Reverse) <- Rarefaction_BT[, 1]
#View(Rarefaction_Bonnethead_Reverse)

sample_data(Rarefaction_BT_Reverse)
summary(Rarefaction_BT_Reverse)
attach(Rarefaction_BT_Reverse)
names(Rarefaction_BT_Reverse)

#ci=2 means that the CI is 95%
sp_BT <- specaccum(Rarefaction_BT_Reverse, object="prey taxa", permutations = 100, ci=2)
#View(sp_BT)

plot(sp_BT, ylab="Number of Prey Taxa", xlab="Number of samples",col="black",ci.type="poly", ci.col="grey", ci.lty = 0,
     xlim=c(1, 110), ylim=c(1, 30)) 
```

Calculate Levin's niche width and Levin's niche overlap - Nursery Site 
```{r}
# Merge data by site name
Gal_Shark_Diet_BT.ps_Site.Name <- merge_samples(Gal_Shark_Diet_BT.ps, group = "Site.Name") 

# Ensure the OTU table has samples as rows and taxa as columns
Gal_Shark_Diet_BT.ps_Site.Name_Reverse <- t(otu_table(Gal_Shark_Diet_BT.ps_Site.Name, taxa_are_rows = TRUE))

# Convert to a data frame
Gal_Shark_Diet_BT.ps_Site.Name_Reverse <- as.data.frame(Gal_Shark_Diet_BT.ps_Site.Name_Reverse)

# View the structure to confirm samples are rows
View(Gal_Shark_Diet_BT.ps_Site.Name_Reverse)

# Calculate niche width for samples
levins.niche.width_species <- niche.width(Gal_Shark_Diet_BT.ps_Site.Name_Reverse, method = "levins")
View(levins.niche.width_species)

# Calculate niche overlap for samples
levins.niche.overlap_species <- niche.overlap(Gal_Shark_Diet_BT.ps_Site.Name_Reverse, method = "levins")
print(levins.niche.overlap_species)
```

Boxplot of Species Richness
```{r}
##Calculate richness by site
# Merge samples by Site.Name
Gal_Shark_Diet_BT.ps_Site <- merge_samples(Gal_Shark_Diet_BT.ps, group = "Site.Name")

richness_data <- estimate_richness(Gal_Shark_Diet_BT.ps_Site, measures = c("Observed"))
View(richness_data)

#Rosa Blanca 2 is potentially lower because half of the samples where taken in 2023 with was El Nino
library(RColorBrewer)
species.richness_island <- plot_richness(
  Gal_Shark_Diet_BT.ps,
  x = "Site.Name",
  color = "Island",
  measures = c("Observed")
) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +  # Slightly transparent boxplot, hide outliers
  geom_jitter(width = 0.2, size = 2, alpha = 0.8) +  # Jitter points for visibility
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_brewer(palette = "Set2")  # Use PRGn palette for coloring by Island

species.richness_island


#####Now calculate with 95% CI
library(phyloseq)
library(dplyr)
library(ggplot2)

# Calculate observed richness from phyloseq object
richness_data <- estimate_richness(Gal_Shark_Diet_BT.ps, measures = "Observed")

# Extract sample metadata (Site Name & Island info)
sample_metadata <- as.data.frame(sample_data(Gal_Shark_Diet_BT.ps))

# Merge richness data with metadata
richness_data$Site.Name <- sample_metadata$Site.Name
richness_data$Island <- sample_metadata$Island

# Compute summary statistics for each site
richness_summary <- richness_data %>%
  group_by(Site.Name) %>%
  summarise(
    mean_richness = mean(Observed, na.rm = TRUE),
    sd_richness = sd(Observed, na.rm = TRUE),
    n = n(),
    se = sd_richness / sqrt(n),  # Standard error
    lower_CI = mean_richness - (1.96 * se),  # 95% CI lower bound
    upper_CI = mean_richness + (1.96 * se)   # 95% CI upper bound
  )

print(richness_summary)  # Check the calculated confidence intervals

# Create base richness plot from phyloseq object
species_richness_plot <- plot_richness(
  Gal_Shark_Diet_BT.ps,
  x = "Site.Name",
  color = "Island",
  measures = c("Observed")
) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +  # Boxplot for spread
  geom_jitter(width = 0.2, size = 2, alpha = 0.8) +  # Jitter points
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_brewer(palette = "Set2")

# Add confidence intervals as error bars
species_richness_plot <- species_richness_plot +
  geom_errorbar(
    data = richness_summary,
    aes(x = Site.Name, ymin = lower_CI, ymax = upper_CI),
    width = 0.2, inherit.aes = FALSE, color = "grey"
  )

print(species_richness_plot)  # Final plot with confidence intervals
```

Richness vs. water temp
```{r}
# Calculate alpha diversity (including Observed taxa richness)
alpha_div <- estimate_richness(Gal_Shark_Diet_BT.ps, measures = c("Observed"))

# Extract sample metadata and convert to a data frame
sample_data_df <- as.data.frame(sample_data(Gal_Shark_Diet_BT.ps))
View(sample_data_df)

# Combine alpha diversity and metadata into a single data frame
combined_data <- cbind(sample_data_df, Observed = alpha_div$Observed)

#Only include samples from Rosa Blanca 2
filtered_data <- combined_data %>%
  filter(Site.Name == "Rosa Blanca 2")

# Filter the data for a specific site (e.g., "Rosa Blanca 2")
Gal_Shark_Diet_filtered.ps <- subset_samples(Gal_Shark_Diet_edited.ps, Site.Name == "Rosa Blanca 2")




library(RColorBrewer)

species.richness_RosaBlanca2 <- plot_richness(
  Gal_Shark_Diet_filtered.ps,
  x = "Date.Collected..MM.DD.YYYY.",
  color = "Water.Temperature..C.",
  measures = c("Observed")
) + 
  geom_boxplot(alpha = 0.6, outlier.shape = NA) + 
  geom_jitter(width = 0.2, size = 2, alpha = 0.8) +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_x_discrete(limits = c("6/23/21", "7/3/21", "7/8/23")) + 
  labs(x = "Date of Collection", y = "Taxa Richness", color = "Water Temp (°C)") + 
  scale_color_distiller(palette = "RdBu", direction = -1)  # Low = blue, High = red

species.richness_RosaBlanca2


#####Now calculate with 95% CI
library(phyloseq)
library(dplyr)
library(ggplot2)

# Combine richness + metadata
richness_data <- estimate_richness(Gal_Shark_Diet_filtered.ps, measures = "Observed")
sample_metadata <- as.data.frame(sample_data(Gal_Shark_Diet_filtered.ps))

# Add date column to richness data
richness_data$Date.Collected <- sample_metadata$Date.Collected..MM.DD.YYYY.
richness_data$Water.Temperature <- sample_metadata$Water.Temperature..C.

# Summarize by collection date
richness_summary <- richness_data %>%
  group_by(Date.Collected) %>%
  summarise(
    mean_richness = mean(Observed, na.rm = TRUE),
    sd_richness = sd(Observed, na.rm = TRUE),
    n = n(),
    se = sd_richness / sqrt(n),
    lower_CI = mean_richness - (1.96 * se),
    upper_CI = mean_richness + (1.96 * se)
  )

# Rename to avoid confusion
species_richness_plot <- species.richness_RosaBlanca2 + 
  geom_errorbar(
    data = richness_summary,
    aes(x = Date.Collected, ymin = lower_CI, ymax = upper_CI),
    width = 0.2,
    inherit.aes = FALSE,
    color = "grey"
  )

print(species_richness_plot)
```

POO Plot separating blacktips by date (only RB2)
```{r}
# Filter the data for a specific site (e.g., "Rosa Blanca 2")
Gal_Shark_Diet_filtered.ps <- subset_samples(Gal_Shark_Diet_edited.ps, Site.Name == "Rosa Blanca 2")

#Convert to presence absence: (FOO)
Gal_Shark_Diet_filtered.ps.PA <- phyloseq_standardize_otu_abundance(Gal_Shark_Diet_filtered.ps, method = "pa")

# Prune taxa with zero abundance across all samples
Gal_Shark_Diet_filtered.ps.PA <- prune_taxa(
  taxa_sums(Gal_Shark_Diet_filtered.ps.PA) > 0,
  Gal_Shark_Diet_filtered.ps.PA)

#Put it into dataframes
otu_df<-cbind(data.frame(otu_table(Gal_Shark_Diet_filtered.ps.PA)))
otu_df$otu<-rownames(otu_df)

otu_df_long<-otu_df%>%
  pivot_longer(cols = -otu,names_to = "sample",values_to = "presence")

otu_df_long$sample <-gsub(pattern = ".",
                          "-",
                          x=otu_df_long$sample,
                          fixed = TRUE)

sample_df<-cbind(data.frame(sample_data(Gal_Shark_Diet_filtered.ps.PA)))
sample_df<-rownames_to_column(sample_df, var = "sample")
sample_date_key<-sample_df%>%
  select(sample,Date.Collected..MM.DD.YYYY.)

taxa_df<-cbind(data.frame(tax_table(Gal_Shark_Diet_filtered.ps.PA)))
taxa_df<-rownames_to_column(taxa_df,var="otu")
otu_species_key<-taxa_df%>%
  select(otu,family,genus,species)

otu_df_long<-otu_df_long%>%
  left_join(sample_date_key)%>%
  left_join(otu_species_key)

#group by different taxonomic levels (shark species)

merge_date_otu_counts<-otu_df_long%>%
  group_by(Date.Collected..MM.DD.YYYY.,species)%>%
  summarise(total_count=sum(presence))%>%
  ungroup()

merge_date_counts<-merge_date_otu_counts%>%
  group_by(Date.Collected..MM.DD.YYYY.)%>%
  mutate(total_site=sum(total_count))%>%
  ungroup()%>%
  mutate(rel_prct=(total_count/total_site)*100)

View(merge_date_counts)

#plot 
POO_BT_date_plot <- ggplot(merge_date_counts, fill = "species") +
  geom_bar(aes(x = Date.Collected..MM.DD.YYYY., y = rel_prct, fill = species), 
           stat = "identity", position = "stack", color = NA) + # Remove borders
  ylab("Relative Percent of Occurrence") + 
  xlab("Date of Collection") +
  theme_classic() + 
  theme(text = element_text(size = 11)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(
    values = c(
      "Abudefduf genus" = "lightblue",
      "Actinopteri class" = "navyblue", "Amphipholis squamata" = "hotpink1", 
      "Anisotremus scapularis" = "palegreen", "Caranx crysos" = "royalblue", 
      "Decapterus genus" = "indianred4", 
      "Euthynnus affinis" = "slategray2", "Fistularia commersonii" = "khaki", 
      "Grapsus grapsus" = "tomato", "Hemiramphus lutkei" = "green3", 
      "Lutjanus genus" = "aquamarine4", "Mugil cephalus" = "blue", 
      "Mugil thoburni" = "purple", "Muraena lentiginosa" = "lightpink", 
      "Pachygrapsus socius" = "goldenrod3", "Sardinops genus" = "red", 
      "Scarus ghobban" = "yellow", "Selar crumenophthalmus" = "slateblue1", 
      "Semicossyphus pulcher" = "lawngreen", "Sphoeroides annulatus" = "lightseagreen", 
      "Sphyrna lewini" = "yellowgreen", "Stegastes arcifrons" = "darkgreen", 
      "Stegastes beebei" = "orange", "Teleophrys cristulipes" = "turquoise2", 
      "Thunnus genus" = "darksalmon"),
    labels = c(
      "Abudefduf genus" = "Abudefduf spp. (sergeant-majors)",
      "Actinopteri class" = "Actinopteri (ray-finned fish)",
      "Amphipholis squamata" = "Amphipholis squamata (dwarf brittle star)",
      "Anisotremus scapularis" = "Anisotremus scapularis (Peruvian grunt)",
      "Caranx crysos" = "Caranx crysos (blue runner)",
      "Decapterus genus" = "Decapterus spp. (jack)",
      "Euthynnus affinis" = "Euthynnus affinis (mackerel tuna)",
      "Fistularia commersonii" = "Fistularia commersonii (bluespotted conetfish)",
      "Grapsus grapsus" = "Grapsus grapsus (Sally lightfoot crab)",
      "Hemiramphus lutkei" = "Hemiramphus lutkei (Lutke's halfbeak)",
      "Lutjanus genus" = "Lutjanus spp. (snapper)",
      "Mugil cephalus" = "Mugil cephalus (flathead grey mullet)",
      "Mugil thoburni" = "Mugil thoburni (Thoburn's mullet)",
      "Muraena lentiginosa" = "Muraena lentiginosa (jewel eel)",
      "Pachygrapsus socius" = "Pachygrapsus Socius (Pacific mottled shore crab)",
      "Sardinops genus" = "Sardinops spp. (sardine)",
      "Scarus ghobban" = "Scarus ghobban (blue-barred parrotfish)",
      "Selar crumenophthalmus" = "Selar crumenophthalmus (bigeye scad)",
      "Semicossyphus pulcher" = "Semicossyphus pulcher (California sheephead)",
      "Sphoeroides annulatus" = "Sphoeroides annulatus (bullseye puffer)",
      "Sphyrna lewini" = "Sphyrna lewini (scalloped hammerhead shark)",
      "Stegastes arcifrons" = "Stegastes arcifrons (yellowtail damselfish)",
      "Stegastes beebei" = "Stegastes beebei (Galapagos ringtail damselfish)",
      "Teleophrys cristulipes" = "Teleophrys cristulipes (spider crab)",
      "Thunnus genus" = "Thunnus spp. (tuna)"))


POO_BT_date_plot
```


